High Resolution Identification of Protein-DNA Binding Events and Quality Control for ChIP-exo data

### 2015-11-16 rmarkdown/chip_exo_paper.Rnw
Replaced saturation, resolution and sensitivity analysis of the paper with new plots using the old data

\usepackage{xcolor}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{float}
\usepackage{verbatim}

@@

\color{blue}

@@

The strand imbalance assesment is based in the observation that the
enriched regions usually have a higher concentration of fragments,
therefore we examine the FSR (defined as the ratio of the number of
forward stranded reads divided by the total number of reads in a given
region) as the region with lower depth are being filtered out. This
indicator is of particular importance, since several methods rely on
the ``peak-pair'' assumption. In table \ref{tab:qcbase}, we calculated
the FSR and noticed that for all the samples, it's value is close to
0.5, which means that there are roughly the same amount of reads in
both strands. To asses

@@

\color{blue}

We next evaluated the ``peak pair'' assumption of Rhee and Pugh
\cite{exo1}, MACE \cite{mace}, CexoR, \cite{cexor} and peakzilla
\cite{peakzilla}, i.e. a peak of reads in the forward strand is
usually paired with a peak of reads in the reverse strand that is
located in the other site of the binding site.

\begin{figure}[h]
  \centering
  \includegraphics[width = .6\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  \caption{ A) ChIP signal in ChIP-exo was linearly related to that in
    PET ChIP-Seq data in the region with high ChIP tag counts. In
    contrast, there wer clear differences in their background
    distribution, where several ChIP-exo background regions were
    empty.}
  \label{fig:comp}
\end{figure}

@@

We sampled a fixed amoung of fragments for each of the ChIP-exo, PET
ChIP-Seq and SET ChIP-Seq datasets of the $\sigma^{70}$ sample in
aerobic conditions. For each sampled dataset we applied our
lower-to-higher resolution pipeline by calling peaks with MOSAiCS
\cite{mosaics} and then deconvoling the binding events by using dPeak
\cite{dpeak}. For the ChIP-exo datasets we called peaks by using
GC-content and mappability with MOSAiCS, and for the ChIP-Seq datasets
we used their respective Input samples.

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/for_paper/Sig70_aerobic_saturation.pdf}
  \caption{Comparison of the number of candidate regions (A),
    predicted events (B), identified targets (C) and resolution (D)
    among ChIP-exo, PET ChIP-Seq and SET ChIP-Seq. RegulonDB
    annotations are considered as a gold standard. A gold standard was
    considered identified if a BS was estimated inside a \Sexpr{ext}
    bp window around it}
  \label{fig:design}
\end{figure}

In figure \ref{fig:design} we compared the behaviour of each data type
when their depth is fixed. It is remarkable that even when the number
of candiate peaks or the number of predicted events in quite lower
than for both ChIP-Seq cases, it outperforms them in the number of
identified targets and in resolution. This may suggest that with
ChIP-exo less false positive peaks are being called and that when the
targets are being identified, dpeak is able to estimate binding
locations closer to the actual true cases. Additionally, we can see
that as the read depth increases all four indicators do so as well,
which may indicate that with ChIP-exo a smaller amount of reads is
necessary to identify a higher number of targets, but it may also be
possible that this is an artifact occurring due to ChIP-exo's lower
library complexity.

@@

\subsection{Method comparison for ChIP-exo}
\label{sec:method_comp}

% \newpage
% \section*{Supplement figures}
% \subsection*{Strand cross-correlation for ChIP-exo data}

@@

\nocite{exo_gb}
\nocite{maplot1}
\nocite{maplot2}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth,page = 3]{../figs/for_paper/Sig70_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page = 4]{../figs/for_paper/Sig70_enrichment.pdf}
  \caption{Hexbin plots of $\mbox{ARC}$ vs $\mbox{URCR}$ for each
    region after partitioning the genome. In A and B the regions with
    reads mapped to at most 10 and 30 positions respectively where not
    considered.}
  \label{fig:enrich2}
\end{figure}

The local strand cross-correlaiton is defined as:

where $f$ is calculated by fitting a non-parametric regression model
over the strand cross-correlation calculated using exclusively the
fragments that overlap a pre-defined genomic region, $y_\delta$ is the
correlation of the forward and backward coverages when one is shifted
$x_\delta$ bp towards their 3' end.

>>>>

\usepackage{xcolor}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{float}
\usepackage{verbatim}
\usepackage{hyperref}

\usepackage{hyperref}
\hypersetup{
     colorlinks   = true,
     citecolor    = black
}
\hypersetup{linkcolor=black}
\hypersetup{linkcolor=black}
\hypersetup{linkcolor=black}

@@

@@

The strand imbalance assesment is based in the observation that the
enriched regions usually have a higher concentration of fragments,
therefore we examine the FSR (defined as the ratio of the number of
forward stranded reads divided by the total number of reads in a given
region) as the region with lower depth are being filtered out. This
indicator is of particular importance, since several methods rely on
the ``peak-pair'' assumption. In table \ref{tab:qcbase}, we calculated
the FSR and noticed that for all the samples, it's value is close to
0.5, which means that there are roughly the same amount of reads in
both strands. However, figure \ref{fig:comp}B shows that this
assumption don't hold in practice.

@@

We next evaluated the ``peak pair'' assumption of Rhee and Pugh
\cite{exo1}, MACE \cite{mace}, CexoR, \cite{cexor} and peakzilla
\cite{peakzilla}, i.e. a peak of reads in the forward strand is
usually paired with a peak of reads in the reverse strand that is
located in the other site of the binding site. In order to evaluate
this assumptiom, we reviewed the proportion of reads in the forward
streand in candidate regions (i.e. regions with at least one binding
site) in $\sigma^{70}$ ChIP-exo data. We found that strands of reads
were much less balanced in ChIP-exo data than in ChIP-Seq data in
these regions with potential binding sites (Fig. \ref{fig:comp}B) and
htis indicates that the peak pair assumptuon might not hold in real
ChIP-exo data.

%  We obtained esentially the same conclusion from the
% ChIP-exo data for CTCF factir from human sample Rhee and Pugh
% (\cite{exo1}) as well.

\begin{figure}[h]
  \centering
  \includegraphics[width = .46\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  \includegraphics[width = .46\textwidth]{../figs/for_paper/forward_strand_ratio_comp_old.pdf}
  \caption{ A) ChIP signal in ChIP-exo was linearly related to that in
    PET ChIP-Seq data in the region with high ChIP tag counts. In
    contrast, there wer clear differences in their background
    distribution, where several ChIP-exo background regions were
    empty. B) Density of the proportion of the ratio of forward strand
    reads in a ChIP-exo and ChIP-Seq SET data.}
  \label{fig:comp}
\end{figure}

\subsubsection{Comparison with ChIP-Seq data using dPeak}
\label{sec:dpeak_analysis}

\begin{figure}[h]
  \centering
  \includegraphics[width = .4\textwidth]{../figs/for_paper/sensitivity_exo_olda_data.pdf}
  \includegraphics[width = .4\textwidth]{../figs/for_paper/resolution_by_dataset_old_data.pdf}
  \caption{Comparison of (A) sensitivity and (B) resolution between
    ChIP-exo and ChIP-Seq data. Sensitivity is defined as the
    propotion of regulonDB annotations identificed using each
    data. Resolution is defined as the distance between regulonDB
    annotation and its closest prediction.}
  \label{fig:reso_all}
\end{figure}

@@

We sampled a fixed amoung of fragments for each of the ChIP-exo, PET
ChIP-Seq and SET ChIP-Seq datasets of the $\sigma^{70}$ sample in
aerobic conditions. For each sampled dataset we applied our
lower-to-higher resolution pipeline by calling peaks with MOSAiCS
\cite{mosaics} and then deconvolving the binding events by using dPeak
\cite{dpeak}. For the ChIP-exo datasets we called peaks by using
GC-content and mappability with MOSAiCS, and for the ChIP-Seq datasets
we used their respective Input samples.

\begin{figure}[h!]
  \centering
%  \includegraphics[width = .8\textwidth]{../figs/for_paper/Sig70_aerobic_saturation.pdf}
  \includegraphics[width = .8\textwidth]{../figs/for_paper/saturation_analysis_old.pdf}
  \caption{Comparison of the number of candidate regions (A),
    predicted events (B), identified targets (C) and resolution (D)
    among ChIP-exo, PET ChIP-Seq and SET ChIP-Seq. RegulonDB
    annotations are considered as a gold standard. A gold standard was
    considered identified if a BS was estimated inside a \Sexpr{ext}
    bp window around it}
  \label{fig:design}
\end{figure}

Figure \ref{fig:design} shows the behaviour of each data type when
their depth is fixed. It is remarkable that even when the number of
candiate peaks or the number of predicted events in quite lower than
for both ChIP-Seq cases, it outperforms them in the number of
identified targets and in resolution. This may suggest that with
ChIP-exo less false positive peaks are being called and that when the
targets are being identified, dpeak is able to estimate binding
locations closer to the actual true cases. Additionally, we can see
that as the read depth increases all four indicators do so as well,
which may indicate that with ChIP-exo a smaller amount of reads is
necessary to identify a higher number of targets, but it may also be
possible that this is an artifact occurring due to ChIP-exo's lower
library complexity.

@@

\subsection{Method comparison with ChIP-exo}
\label{sec:suppcomp}

We considered dPeak Chung et al. \cite{dpeak}, GEM Guo et
al. \cite{gem} and MACE Wang et al. \cite{mace} for ChIP-exo data
analysis. For the dPeak algorithm, we used the R package \emph{dPeak}
version 1.0.0, which is available from
\url{http://dongjunchung.github.io/dpeak/}. For the GEM algorithm we
used its Java implementation version 0.9, which is available from
\url{http://cgs.csail.mit.edu/gem/}. For the MACE algorithm, we used
ity Python implementation version 1.0, which is available from
\url{http://dldcc-web.brc.bcm.edu/lilab/MACE/docs/html/}. Candidate
regions for dPeak were identified for each replicate of ChIP-exo data
using the MOSAiCS algorithm Kuan et al. \cite{mosaics} (one sample
analysis using false discovery rate of 0.0001) implemented as an R
package \emph{mosaics} 2.4 (available from \emph{Bioconductor}
\url{https://www.bioconductor.org/packages/release/bioc/html/mosaics.html}). We
further filtered out candidate region with average ChIP tag count less
than 3,000 to avoid potential false positives based in exploratory
analysis. These regions were also explicitly provided to the GEM
algorithm as candidate regions. Default tuning parameters were used
duging model fitting for all methods.

The local strand cross-correlation is defined as:

where $f$ is calculated by fitting a local polynomial regression model
over the strand cross-correlation calculated using exclusively the
fragments that overlap a pre-defined genomic region, $y_\delta$ is the
correlation of the forward and backward coverages when one is shifted
$x_\delta$ bp towards their 3' end.

The local polynomial model was fitted using the \emph{loess} function
from the R software version 3.2.1 using the default tunning parameters.

%% not sure if here or before it would be nice to add an example of how this 
%% method works. probably here

@@

\nocite{exo_gb}
\nocite{maplot1}
\nocite{maplot2}
% \nocite{esl}

@@

In figure \ref{fig:enrich2} it is shown the same relationshtip as in
figure \ref{fig:enrich}, but considering only regions formed where the
reads are being allocated in more than 10 (A) and 30 (B) unique
positions respectively. Both plots show that the vertical arms formed
by regions with low $\mbox{ARC}$ is formed by low complexity regions,
that way suggesting that this segment correspond to the background of
a ChIP-exo experiment.

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth,page = 3]{../figs/for_paper/Sig70_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page = 4]{../figs/for_paper/Sig70_enrichment.pdf}
  \caption{Hexbin plots of $\mbox{ARC}$ vs $\mbox{URCR}$ for each
    region after partitioning the genome. In A and B the regions with
    reads mapped to at most 10 and 30 positions respectively where not
    considered. $\mbox{ARC}$ is defined as the ratio of the nr. of
    reads and the width of a region and $\mbox{URCR}$ is the ratio of
    the number of unique position where the reads are being allocated
    and the number of reads in a region.}
  \label{fig:enrich2}
\end{figure}

### 2015-11-16 rmarkdown/chip_exo_paper.tex
Replaced saturation, resolution and sensitivity analysis of the paper with new plots using the old data

We sampled a fixed amoung of fragments for each of the ChIP-exo, PET
ChIP-Seq and SET ChIP-Seq datasets of the $\sigma^{70}$ sample in
aerobic conditions. For each sampled dataset we applied our
lower-to-higher resolution pipeline by calling peaks with MOSAiCS
\cite{mosaics} and then deconvoling the binding events by using dPeak
\cite{dpeak}. For the ChIP-exo datasets we called peaks by using
GC-content and mappability with MOSAiCS, and for the ChIP-Seq datasets
we used their respective Input samples.

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/for_paper/Sig70_aerobic_saturation.pdf}
  \caption{Comparison of the number of candidate regions (A),
    predicted events (B), identified targets (C) and resolution (D)
    among ChIP-exo, PET ChIP-Seq and SET ChIP-Seq. RegulonDB
    annotations are considered as a gold standard. A gold standard was
    considered identified if a BS was estimated inside a 15
    bp window around it}
  \label{fig:design}
\end{figure}

In figure \ref{fig:design} we compared the behaviour of each data type
when their depth is fixed. It is remarkable that even when the number
of candiate peaks or the number of predicted events in quite lower
than for both ChIP-Seq cases, it outperforms them in the number of
identified targets and in resolution. This may suggest that with
ChIP-exo less false positive peaks are being called and that when the
targets are being identified, dpeak is able to estimate binding
locations closer to the actual true cases. Additionally, we can see
that as the read depth increases all four indicators do so as well,
which may indicate that with ChIP-exo a smaller amount of reads is
necessary to identify a higher number of targets, but it may also be
possible that this is an artifact occurring due to ChIP-exo's lower
library complexity.

>>>>

We sampled a fixed amoung of fragments for each of the ChIP-exo, PET
ChIP-Seq and SET ChIP-Seq datasets of the $\sigma^{70}$ sample in
aerobic conditions. For each sampled dataset we applied our
lower-to-higher resolution pipeline by calling peaks with MOSAiCS
\cite{mosaics} and then deconvolving the binding events by using dPeak
\cite{dpeak}. For the ChIP-exo datasets we called peaks by using
GC-content and mappability with MOSAiCS, and for the ChIP-Seq datasets
we used their respective Input samples.

\begin{figure}[h!]
  \centering
%  \includegraphics[width = .8\textwidth]{../figs/for_paper/Sig70_aerobic_saturation.pdf}
  \includegraphics[width = .8\textwidth]{../figs/for_paper/saturation_analysis_old.pdf}
  \caption{Comparison of the number of candidate regions (A),
    predicted events (B), identified targets (C) and resolution (D)
    among ChIP-exo, PET ChIP-Seq and SET ChIP-Seq. RegulonDB
    annotations are considered as a gold standard. A gold standard was
    considered identified if a BS was estimated inside a 15
    bp window around it}
  \label{fig:design}
\end{figure}

Figure \ref{fig:design} shows the behaviour of each data type when
their depth is fixed. It is remarkable that even when the number of
candiate peaks or the number of predicted events in quite lower than
for both ChIP-Seq cases, it outperforms them in the number of
identified targets and in resolution. This may suggest that with
ChIP-exo less false positive peaks are being called and that when the
targets are being identified, dpeak is able to estimate binding
locations closer to the actual true cases. Additionally, we can see
that as the read depth increases all four indicators do so as well,
which may indicate that with ChIP-exo a smaller amount of reads is
necessary to identify a higher number of targets, but it may also be
possible that this is an artifact occurring due to ChIP-exo's lower
library complexity.
